\documentclass[a4paper]{article}

\usepackage{hyperref}
\usepackage{xurl}

\title{petope - a VPN development project}
\author{Retro}
\date{2026}

\begin{document}
  \maketitle

  \tableofcontents

  \addcontentsline{toc}{section}{Introduction}
  \section*{Introduction}

  When someone asks how can they connect to their server at home, 
  for example on \href{https://www.reddit.com/r/selfhosted/}{r/selfhosted} subreddit,
  I often see that \href{https://tailscale.com/}{Tailscale} is suggested.

  Why you may ask? Well to connect to your own server is not as easy task as 
  to connect normal public facing server. 
  Most of home networks are behind NAT with periodically changing IPv4 address
  \href{https://en.wikipedia.org/wiki/IPv4_address_exhaustion}{since we don't have enough IPv4 addresses},
  and there are firewalls that prevent access from outside.
  
  In some cases you don't even have public IPv4 address, 
  since ISP provided may use \href{https://en.wikipedia.org/wiki/Carrier-grade_NAT}{CGNAT} 
  which makes connection from the internet impossible even if you use \href{https://en.wikipedia.org/wiki/Dynamic_DNS}{Dynamic DNS}
  and open ports on your router.

  And most of these problems are solved by Tailscale! 
  When you start a Tailscale application, it connects to a \href{https://tailscale.com/docs/concepts/control-data-planes}{control plane}
  which coordinates your Tailscale network (known as \href{https://tailscale.com/docs/concepts/tailnet}{tailnet}),
  tells you which devices are in it, their internal addresses, public key, manages access to each device
  and helps you with \href{https://tailscale.com/blog/how-nat-traversal-works}{NAT traversal} 
  so you can even connect directly to your server.
  Instead of classic VPN structure where you connect to relay server to access internal subnet
  you connect to each of your servers directly, thus cutting a middle man and an additional latency
  and single point of failure\footnote{there still a single point of failure, as tailnet is managed by control plane instead of relay server}.
  And since all clients in the network connect to each other directly, 
  it is called a \href{https://en.wikipedia.org/wiki/Mesh_networking}{Mesh VPN}.

  So, with Tailscale you just connect your device to your tailnet, and you can just to your other devices by an address. 
  No need to add peers to each device, open ports, manage routing. Plug-n-play solution that just works.

  Also it makes a great solution when you want to play with your friends together,
  but there is much simpler to use alternative named \href{https://www.radmin-vpn.com/}{Radmin VPN},
  which has somewhat similiar functionality, but is proprietary, and I just don't like it.

  So, after being amazed by how Tailscale works internally, I want to learn by
  creating my own VPN solution, learn from my mistakes and while developing everything needed,
  and maybe even create something for gamers around the world to use to play together.
  And I am totally okay if I won't be able to handle such project, I am sure I will learn a lot of stuff!


  \section{Researching existing solutions} \label{existingsolutions}

  Since I like \href{https://en.wikipedia.org/wiki/Self-hosting_(network)}{self hosting}
  different stuff on my servers, I also like to have control of my data,
  particulary don't give a third-party control to it. 
  That's why I don't fully enjoy Tailscale and Radmin VPN solutions:
  \begin{itemize}
    \item Radmin VPN is fully proprietary, so they control the server- and client-side.
          There is nothing stopping them from adding collecting data or even stealing from my PC,
          or adding unauthorized device to my network.
    \item Tailscale has open-source client, but control-plane is fully proprietary and controls
          how client operates. In case of its compomise, your connected devices are fully open
          unless you have configured a firewall. But I trust Tailscale company much more,
          and it feels like a fair trade of security and convinience.
  \end{itemize}

  But projects like \href{https://syncthing.net/}{Syncthing} and \href{https://yggdrasil-network.github.io/}{Yggdrasil}
  inspire me to create something more decentralized, where each device in control of to whom to trust,
  and don't rely on single server which commands to whom he can connect and how.

  I really liked Yggdrasil approach to the network -
  each client is also a server, and an IPv6 address of a client also represents his public key.
  This way when you send a packet to an IPv6 address, you can already encrypt message
  by using the address, and only client who has right private key can decrypt it.
  It is very cool idea since without a single authority you can be sure
  that only correct recipient can decrypt and see your messages, and other peers
  in the network relay the message until it gets to the right recipient.
  Decentralization power, hell yeah!

  But without a central server there is no way to find a way how to connect to a peer on the internet. 
  \href{https://en.wikipedia.org/wiki/BitTorrent}{BitTorrent}
  % https://en.wikipedia.org/wiki/Peer_exchange
  % https://en.wikipedia.org/wiki/Mainline_DHT


  \section{Planning} \label{planning}

  Before starting any implementation, 
  I need to decide important choices in 
  how I will be developing my own VPN:
  \begin{enumerate}
    \item \nameref{corefunctionality}
    \item 
  \end{enumerate}

  \subsection{Core functionality} \label{corefunctionality}

 
  \newpage
  \section*{Sources}
  \begin{itemize}
    \item \url{https://yggdrasil-network.github.io/implementation.html}
    \item \url{https://yggdrasil-network.github.io/2018/08/19/congestion-control.html}
    \item \url{https://yggdrasil-network.github.io/2018/07/13/about-mtu.html}
    \item \url{https://yggdrasil-network.github.io/2023/10/22/upcoming-v05-release.html}
  \end{itemize}

  % $ https://www.reddit.com/r/crypto/comments/f7c2nv/chacha20_v_aes256/
  % $ https://crypto.stackexchange.com/questions/79518/is-xchacha20-poly1305-quantum-resistant
  % $ https://crypto.stackexchange.com/questions/101050/xchacha20-vs-aes-128-security-and-speed
  % https://nordpass.com/features/xchacha20-encryption/

\end{document}
